master thesis

### 2018-03-28 Paper - Latex/Paper - Skizze - v4.tex
Thesis - Introduction; Methods bis real data analysis

  Comparing the methods, we find that parsimony are methods in which you can not include information 
    such as branch length (development time of the species). This is also their main disadvantage 
    compared to the more complex likelihood based models. Since there are no branch lengths or other 
    information available for the present relationship tree, and we can thus not take advantage of 
    the other models. \\
  The comparison of methods gives us the maximum parsimony method as the simplest and fastest method 
    that meets all our requirements.

@@

  The Sankoff method is implemented by Louca et al. for the non-binary case and is available as an R 
    package called \textit{Castor} \cite{Louca2017}. In addition, we have implemented the Fitch 
    method \cite{Fitch1971} \todo{hier vll die andere quelle nutzen nach der wir uns gerichtet haben?}and adapted it for multifurcated trees. \\% to make it usable and comparable for our current problem.
  This achieves that we can predict ancestral states and unknown states of living species for large 
    non-binary trees.

  To accomplish this task, next to the phylogenetic tree information about the current species' 
    states is needed. \\
  For this information we use the interaction database Global Biotic Interactions (GloBI) 
    \cite{Poelen2014}. The data in GloBI are stored as interactions e.g.  species A parasitize 
    species B. From this we conclude that species A is parasitic and species B free-living. \\
  From this data, we could specify $\sim$2.3 million leaf nodes 34,879 as free-living and 25,992 
    as parasitic.

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Methods}
  As stated in the introduction, in this thesis, a maximum parsimony algorithm is applied to the 
    whole tree of life to obtain an ancestral state reconstruction of free-living versus parasite 
    states. \\
  So far, these reconstructions have been made mainly on binary trees with better data availability. 
    Therefore, a simulation is first performed to evaluate existing algorithms and decide how they 
    may be adapted to our given problem. This is to perform the ancestral state reconstruction for a 
    multifurcated (non-binary) tree using binary states. \\
  Accordingly, in addition to the necessary data sets (GloBI, OTL), the chosen algorithm and the 
    evaluation of its results, this chapter also deals with the previously performed simulation and 
    the evaluation of the various algorithms and their parameters. \\
  Figure \ref{fig:workflow} briefly outlines these relationships. A more detailed view of the 
    workflow can be found in the appendix \ref{fig:BigWorkflow}.
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/Workflow-overview.png}
    \caption{The Workflow of the resulting procedure with the following steps: \\
      (1) Retrieve phylogenentic tree data as input for the tree (OTL) and the state information (GloBI).
      (2) Get metadata of these for a realisitc simulation of the maximum parsimony algorithms (Fitch \& Sankoff).
      (3) Build and run the simulation.
      (4) Evaluation of parameters for the simulation and the ancestral state reconstruction of the real tree.
      (5) Evaluate the accuracy of developed algorithms and choose the best.
      (6) Run the resulting algorithm on the original data.
      (7) Evalute and interprete the results. %$\rightarrow$ Origins etc...
    }
    \label{fig:workflow}
  \end{figure}

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Description of data sets}
    Two types of data are needed for an ancestral state reconstruction: a tree and information about 
      the states. \\
    For the tree Open Tree of Life (OTL) is used \cite{Hinchliff2015} and for the state information 
      the Global Biotic Interactions database (GloBI) is taken \cite{Poelen2014}. \\
    % These are presented below.

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{OTL}
      For this project a large database for phylogenetic trees and also for a taxonomic tree is needed. 
        Since an ancestral state reconstruction algorithm is applied to the phylogenetic tree, and for 
        the assessment and other properties the taxonomy provides much more information. \\
      OTL gives a synthesis of phylogenetic trees (currently 819 trees) and a taxonomic tree\footnote{
          \hyperlink{https://tree.opentreeoflife.org/about/synthesis-release/v9.1}
          {https://tree.opentreeoflife.org/about/synthesis-release/v9.1}; Last checked: 22.03.2018.
        }. OTL also 
        includes the large phylogenetic database TreeBASE \cite{Hinchliff2015}. \\
      \todo{Das steht auf der Website nicht in dem Paper...} \\
      % For taxonomic data, there are at least five big data collections, namely:
      % \begin{itemize}
      %   \item ITIS (Integrated Taxonomic Information System) \cite{ITIS}
      %   \item NCBI (National Center for Biotechnology Information) \cite{NCBI1988}
      %   \item WORMS (World Register of Marine Species) \cite{WoRMS2018}
      %   \item GBIF (Global Biodiversity Information Facility) \cite{GBIF}
      %   \item OTT (OpenTreeOfLife-Taxonomy) \cite{Hinchliff2015}
      % \end{itemize}
      % ITIS is only a small set of 100\% confirmed and named species. GBIF is not composed with the help 
      %   of phylogeny, the same is valid for the NCBI taxonomy. The WORMS taxonomy is a way too small 
      %   dataset of mostly marine species. \\
      Here the taxonomy from OTL is used because it is including most of the known taxonomies and is 
        synthesised by preffering taxonomies that match with available phylogenetic data. Furthermore 
        the team from OTL preferre a maximum number of species \cite{Hinchliff2015}. This is resulting 
        in somekind of hybrid between taxonomy and phylogeny. \anmerkungstext{Wie genau ist das ein 
        Hybrid? Genauer beschreiben, was Du damit meinst... (Thilo)} \\

      A closer look is being made to some of the features of the synthesis tree. On the one hand the 
        distribution of the taxa and on the other the distribution of the nodes on the taxa. Since this 
        is not directly relevant for this study, there is a section in the appendix 
        \ref{sec:otl analysis}.

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{GloBI}
      The most big interaction databases are offline or outdated. For example: IWDB (Interaction Web 
        Database) \cite{IWDB2003}, Webs on the Web \cite{WOW2004}, Animal Diversity Web \cite{Myers2003} 
        and ecoweb \cite{Cohen2010}. GloBI is including most of the known ones and is still growing 
        actively \cite{Poelen2014}. So the question was answered rather quickly which interaction 
        database could be used. \\

      This database consists of entries of the form: species A (source) interacts with B (target). \\
      A number of interactions have been identified\footnote{\hyperlink{
        https://github.com/jhpoelen/eol-globi-data/blob/master/eol-globi-lib/src/main/java/org/eol/globi/domain/InteractType.java
        }{https://github.com/jhpoelen/eol-globi-data/.../InteractType.java}; Last checked: 22.03.2018.}
        , including those that the species source or target has become a parasite or a free-living 
        species from the biological perspective. These are the following:
      \begin{itemize}
        \item free-living source: preysOn, eats, flowersVisitedBy, hasPathogen, pollinatedBy, 
          hasParasite, hostOf
        \item free-living target: preyedUponBy, parasiteOf, visitsFlowersOf, pathogenOf, hasHost
        \item parasite source: parasiteOf, pathogenOf
        \item parasite target: hasParasite, hasPathogen
      \end{itemize}
      Of these interactions, e.g. species A parasitize species B, the state of the species is 
        determined, here is species A parasitic and species B free-living. The case a parasite 
        conquers (parasitizes) another parasite yields conflicting states for the second species. 
        This is solved by preferring parasitic. \\
      For each species known IDs are stored in GloBI. This includes OTT (the taxonomy of OTL). All 
        species that have stored an OTT id and have a matching interaction are formed into two lists: 
        parasites and free-livings.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Metadata analysis}
    In order to generate a realistic simulation, influencing parameters are investigated. \\
    Since the transitions are minimized in an ancestral state reconstruction, this is an important 
      parameter to consider. On the other hand, the completeness of our input data are influencing 
      values. Therefore, two major types are distinguished:
    \begin{enumerate}
      \item Biological parameters (A result of the evolutionary process.):
        \begin{itemize}
          \item transition probabilities
        \end{itemize}
      \item Distribution of the loss of information:
        \begin{itemize}
          \item Loss of topology ($\rightarrow$ mutlifurcations)
          \item Unknown information about states of some leaf nodes
        \end{itemize}
    \end{enumerate}
    The influence of these parameters are tested on our result using our simulation (section 
      \ref{sec:methods - simulation}).

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Transition probabilities}
      As mentioned above, in an ancestral state reconstruction, transitions are minimized, with 
        transition probabilities playing a role. \\
      This subsection deals with these transition probabilities from free-living (hereinafter / as a 
        formula FL) to parasitic (hereinafter P) and vice versa: $\mathcal{P}(FL \rightarrow P)$, 
        $\mathcal{P}(P \rightarrow FL)$. \\
      Different parasite types have different transition probabilities. It is very difficult to make a 
        statement about these probabilities. It is generally assumed in this work that there are 40\% 
        parasites and 60\% free-livings which is based on the estimates by Windsor \cite{Windsor1998} 
        and $\mathcal{P}(FL \rightarrow P) = \mathcal{P}(P \rightarrow FL)$, as a result of the 
        diversity of parasites and the lack of general determinations for this. These parameters are 
        debated in subsection \ref{sec:discussion - transition probabilities} of the discussion. \\

      For the maximum parsimony analysis of the real data, all transition probabilities are equated.
        However, the used castor package \cite{Louca2017} offers the possibility to enter different 
        transition probabilities.
      \begin{wrapfigure}{r}{0.5\textwidth}
        \begin{center}
          \includegraphics[trim = 0mm 0mm 0mm 0mm, clip, width=0.45\textwidth]{Figures/40-60.png}
        \end{center}
        \caption{Simulated distributions of the states: \\
          60\% Free-living - 40\% Parasites \\
          red: parasites, blue: free-living, \\
          the threshold is at 0.4}
        \label{fig:Beta distribution}
      \end{wrapfigure}

      In the simulation two beta distributions have been chosen and a threshold that indicates the change 
        between states. \\
      Different thresholds with different beta distributions are simulated, with different distributions 
        of parasites and free-livings:
        \begin{itemize}
          \item 50\% P to 50\% FL,
          \item 40\% P to 60\% FL,
          \item 30\% P to 70\% FL and 
          \item 20\% P to 80\% FL
        \end{itemize}
        (\todo{see results simulation, ref...}). Figure \ref{fig:Beta distribution} 
        shows one example of these. \\

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Missing information} \label{sec:methods - missing information}

      A binary tree with $n$ leaf nodes has $n-1$ internal nodes. The present Eukaryota tree of OTL has 
        2,293,463 leaf nodes and only 41,974 internal nodes, that is:
      $$100-\frac{100}{(2293463-1) \times 41974} \approx 98.16 \%$$
        missing internal nodes. This means that there is a lack of information about the underlying 
        phylogeny. Instead of a binary tree this tree is highly multifurcated. \\

      For the present Eukaryota tree with 2,293,463 leaf nodes, 34,869 free-livings and 25,962 parasites 
        are found, which are
        $$100-\frac{100}{2293463 \times (34860+25962)} \approx 97.34 \%$$
        unknown states of leaf nodes. \\

      In the simulation, the influence of the multifurcation and missing data in leaf nodes on the 
        predictive accuracy of the ancestral state reconstruction algorithms is tested. \\
      For the real data, generalized linear models are compared with poisson respectively binomial 
        regression according to their residuals. In order to compare models of different complexity, the 
        BIC (Bayesian Information Criterion) values were calculated in addition to the residuals.\\
      There are two different information criteria: AIC (Akaike Information Criteria) and BIC. \\
      The advantage of the BIC is that the penalty is dependent on the sample size and is therefore 
        advantageous for large samples. \\

      For each node, depth, min, max and mean height were noted. Where the depth of a node is the 
        distance (number of edges) to the root node and the height of a knot is described as the 
        largest distance to a leaf node. In this work, a distinction is made between minimum, maximum 
        and average distance ($\rightarrow$ min, max and mean height). \\
      The influence in the modeling of these parameters was tested, additive as well as multiplicative. \\

      For all these calculations, the following R functions were used: \textit{glm()}, \textit{anova()} 
        and \textit{BIC()}.
      % https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Ancestral state reconstruction methods} \label{sec:ancestral state reconstruction}
    As presented in the introduction, there are some methods for ancestral state reconstruction. For 
      this purpose, various studied methods and their advantages and disadvantages are compared below. \\

    Royer-Carenzi et al. distinguishes two major classes of ancestral state reconstruction methods: \\
    The first is to explain the current state with the least number of state changes between an 
      ancestor and his child, this is called parsimonious. \\
    The other class she presents involves modeling the character evolution as a stochastic process and 
      using the likelihoods to compute the possible ancestral character states. This is generally done 
      with a continuous time Markov model \cite{RoyerCarenzi2013}. \\

    One of the major disadvantages of parsimony methods is that, unlike likelihood approaches, they 
      can not take divergece times (branch length) into account. Since the OTL does not include 
      development times of species, this is irrelevant. \\
    Another problem pointed out by Royer-Carenzi is that parsimony approaches are either based on 
      predefined parameters (generalized parsimony) or on strong and often controversial assumptions, 
      like irreversibility of transitions for dollo parsimony. Again, this problem is unimportant to 
      the problem at hand, because in the analysis of the entire Eukaryota tree only generalized 
      models make sense. \\

    Parsimony-based methods are used in this work, since they are fully sufficient for the presented 
      use case here. Following the principle of the simpler model first. \\
    \anmerkungstext{Nochmal darstellen warum und selbst wenn sie fully sufficient wäre, heißt es ja 
      nicht, dass man sie nehmen muss. Also: motiviere! (Renard)} \\
    Felsenstein \cite{Felsenstein2003} discusses in his book two algorithms that generalize all 
      previous methods (from Camin and Sokal \cite{Camin1965}, \todo{Kluge and Farris} and Farris 
      \cite{Farris1970}): Fitch parsimony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. \\
    \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
      Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
    \todo{Wagner-parsimony \cite{Swofford1987}} \\

    Thus, the methods used in this work are those of Fitch and Sankoff. For Fitch, the algorithm has 
      been extended from binary to multifurcated trees. For the Sankoff algorithm, Louca and Doebeli 
      have presented an implementation for non-binary trees published in an R package named 
      \textit{castor} \cite{Louca2017}.

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Fitch maximum parsimony}
      Fitch maximum parsimony is an algorithm for rooted, binary trees and describes an ancestral state 
        reconstruction for discrete states \cite{Fitch1971} by minimizing transitions between states. \\
      Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
        and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
        simple extension for the reconstruction. Cunningham et al. \cite{Cunningham1998} have refined 
        these. \todotext{Wir haben mit ein paar kleinen änderungen optimiert... und schließlich auf 
        multifurcated angepasst...} \todotext{eigentlich ist Cunningham 'nur' eine kritische 
        Neubewertung. Sie beziehen ihren Algortihmus auf Swofford und Maddison...} \\
      To understand the differences to the multifurcated case, the algorithm for the binary case is 
        briefly explained and referred to the extension. \\

      Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (e.g. \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}). \\

      The algorithm traverses three times through the tree and fills these sets. \\
      In each step, two sets are considered and their intersect formed. There are two cases:
      \begin{enumerate}
        \item The intersection is not empty and corresponds to the new set.
        \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
      \end{enumerate}
      \begin{wrapfigure}{r}{0.5\textwidth}
        \centering
        \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
        \caption{Fitch algorithm for binary trees. \\
          The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
          root node) consists of three sets, where the last set is the final one (bold). \\
          From the second internal node (seen from the root node) there are several possibilities to 
          create the second and third set.}
        \label{fig: binary Fitch}
      \end{wrapfigure}
      First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.

        Each internal node is formed from its child nodes, where at the beginning the only information
        lies. \\
      Second traverse from the root node to the leafs. Each internal node is formed from its father node 
        and its sibling node. \\
      Last traversion (direction does not matter): Build the final state for every node. It is formed 
        from the sets of previous traversals. \\
      (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
        of internal nodes, so it was just the first traversal.) \\
      The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
        more than two children may be present for the first traversal, but the incision or union may 
        also be formed over more than two sets. Also in the second traversing, there may be several 
        sibling nodes. However, there are several possibilities here that were all tested and compared 
        in the simulation. Some of these options are already available in the binary case:
      \begin{itemize}
        \item The father node has (except for the root node) two state sets, because he came through 
          the up-traversing previously. Are both sets used or only the first traversing?
        \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
          or directly in the whole with the father node?
      \end{itemize}
      The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
        both possibilities of the three sets. \\
      Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
      From these two points four different versions of Fitch were formed:
      \begin{enumerate}
        \item Fitch 1: First state set of father node; intersection/union of siblings first.
        \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
        \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
        \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
      \end{enumerate}
      \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
        \caption{Test of Fitch Versions.}
        \label{fig:Fitch versions}
      \end{figure}
      These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
        distribution of 60\% FL to 40\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
        node percentage. \\
      % At 90\% unknown nodes and 90\% of multifurcation of the internal nodes, version 1 was 89.26\%, 
      %   version 2 was 89.26\%, version 3 was 89.35\%, and version 4 was 89.31\% correct. Therefore, 
      %   only version 3 was used for all further simulations.
      At 95\% unknown nodes and 95\% of multifurcation of the internal nodes, version 1 was 88.37\%, 
        version 2 was 88.37\%, version 3 was 88.4\%, and version 4 was 88.39\% correct. Therefore, 
        only version 3 was used for all further simulations.

@@

    %---------------------------------------------------------------------------------------------------
    \subsubsection{Sankoff}
      Maximum parsimony algorithm from Sankoff implemented in the R package castor \cite{Louca2017}. \\
      \todo{transition probabilities: all equal}

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Simulation}\label{sec:methods - simulation}
    The simulation compares these different of ancestral state reconstruction algorithms with each 
      other. \\
    First different implementations of the Fitch maximum parsimony are compared and then the best of 
      them is compared with the implementation of the Sankoff algorithm of the \textit{Castor} package 
      \cite{Louca2017}. \\

>>>>

  One of the major disadvantages of parsimony methods is that, unlike likelihood approaches, they 
    can not take divergence times (branch length) into account. Since the OTL does not include 
    development times of species, this can not be considered here. \\
  Another problem pointed out by Royer-Carenzi is that parsimony approaches are either based on 
    predefined parameters (generalized parsimony) or on strong and often controversial assumptions, 
    like irreversibility of transitions for dollo parsimony. This problem is unimportant to the 
    present task because in the analysis of the entire Eukaryota tree only generalized models are 
    meaningful. \\
  % Comparing the methods, we find that parsimony are methods in which you can not include information 
  %   such as branch length (development time of the species). This is also their main disadvantage 
  %   compared to the more complex likelihood based models. Since there are no branch lengths or other 
  %   information available for the present relationship tree, and we can thus not take advantage of 
  %   the other models. \\
  The comparison of methods gives us the maximum parsimony method as the simplest and fastest method 
    that meets all our requirements.
  % Parsimony-based methods are used in this work, since they are fast and fully sufficient for the 
  %   presented use case here, following the principle of the simpler model first. \\

  Felsenstein \cite{Felsenstein2003} discusses in his book two parsimony algorithms that generalize 
    previous methods (from Camin and Sokal \cite{Camin1965}, Farris \cite{Farris1970} and others): 
    Fitch parsimony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. Therefore, these are 
    the methods used in this work. \\
  % \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
  %   Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
  % \todo{Wagner-parsimony \cite{Swofford1987}} \\
   For Fitch, the algorithm has been extended from binary to multifurcated trees. For the Sankoff 
    algorithm, Louca and Doebeli have presented an implementation for non-binary trees published in 
    an R package named \textit{castor} \cite{Louca2017}.

@@

  The Sankoff method is implemented by Louca et al. for the non-binary case and is available as an R 
    package called \textit{Castor} \cite{Louca2017}. \\
  Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
    and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
    simple extension for the reconstruction. In this work, the algorithm extended to reconstruction 
    is adapted to multifurcated trees, based on the critical reevaluation of this extension by 
    Cunningham et al. \cite{Cunningham1998}.  
  % In addition, we have implemented the Fitch 
  %   method \cite{Fitch1971} \todo{hier vll die andere quelle nutzen nach der wir uns gerichtet haben?}and adapted it for multifurcated trees. \\% to make it usable and comparable for our current problem.
  % This achieves that we can predict ancestral states and unknown states of living species for large 
  %   non-binary trees.

  To accomplish this task, next to the phylogenetic tree information about the current species' 
    states is needed. \\
  Most of the largest interaction databases are offline or outdated. For example: IWDB (Interaction 
    Web Database) \cite{IWDB2003}, Webs on the Web \cite{WOW2004}, Animal Diversity Web 
    \cite{Myers2003} and ecoweb \cite{Cohen2010}. \\
  We use the interaction database Global Biotic Interactions (GloBI) \cite{Poelen2014} because it is 
    including most of the known databases and is still growing actively \cite{Poelen2014}. \\
  The data in GloBI are stored as interactions e.g.  species A parasitize species B. From this we 
    conclude that species A is parasitic and species B free-living. \\
  From this data, we could specify $\sim$2.3 million leaf nodes 34,879 as free-living and 25,992 
    as parasitic.

@@

  In order to generate a realistic simulation, influencing parameters are investigated. \\
  Since the transitions are minimized in an ancestral state reconstruction, this is an important 
    parameter to consider. On the other hand, the completeness of our input data is an influencing 
    value. Therefore, two major types are distinguished:
  \begin{enumerate}
    \item Biological parameters (a result of the evolutionary process):
      \begin{itemize}
        \item State distributions and transition probabilities
      \end{itemize}
    \item Distribution of missing information:
      \begin{itemize}
        \item Lack of information on topology ($\rightarrow$ multifurcations)
        \item Lack of information of states of some leaf nodes
      \end{itemize}
  \end{enumerate}

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Methods}
  In this thesis, a maximum parsimony algorithm is applied to the Eukaryota tree to obtain an 
    ancestral state reconstruction of free-living versus parasite states. \\
  This chapter is divided into the following sections: the description of the data sources 
    (section \ref{sec:methods - data sources}) used and the analysis of these data as a preparation 
    for the simulation (section \ref{sec:methods - metadata analysis}). A description of the 
    analyzed methods for the ancestral state reconstruction (section 
    \ref{sec:methods - ancestral state reconstruction}) and then an explanation of how they fit the 
    problem at hand (section \ref{sec:methods - simulation}). And at the end, the real data analysis 
    with the Sankoff method (section \ref{sec:methods - real data analysis}). \\
  Figure \ref{fig:workflow} briefly outlines these relationships. A more detailed view of the 
    workflow can be found in the appendix \ref{fig:BigWorkflow}.
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/Workflow-overview.png}
    \caption{The Workflow of the resulting procedure with the following steps: \\
      (1) Retrieve phylogenentic tree data as input for the tree (OTL) and the state information (GloBI).
      (2) Get metadata of these for a realisitc simulation of the maximum parsimony algorithms (Fitch \& Sankoff).
      (3) Build and run the simulation.
      (4) Evaluation of parameters for the simulation and the ancestral state reconstruction of the real tree.
      (5) Evaluate the accuracy of developed algorithms and choose the best.
      (6) Run the resulting algorithm on the original data.
      (7) Evalute and interprete the results. %$\rightarrow$ Origins etc...
    }
    \label{fig:workflow}
  \end{figure}

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Data sources} \label{sec:methods - data sources}
    Two types of data are needed for an ancestral state reconstruction: a tree and information about 
      the states.

    For the subsequent analyzes on the Eukaryote tree the database of Open Tree of Life (OTL) was used 
      (downloaded on 16.02.2018) \cite{Hinchliff2015}. This database gives a synthesis of phylogenetic 
      trees (currently 819 trees) and a taxonomic tree\footnote{
        \hyperlink{https://tree.opentreeoflife.org/about/synthesis-release/v9.1}
        {https://tree.opentreeoflife.org/about/synthesis-release/v9.1}; Last checked: 22.03.2018.
      }. OTL also  includes the large phylogenetic database TreeBASE \cite{Hinchliff2015}.

    Furthermore the Open Tree Taxonomy (OTT) from OTL is used because it includes most of the known 
      taxonomies and is synthesised by prefering taxonomies that match with available phylogenetic 
      data. The team from OTL prefer a maximum number of species \cite{Hinchliff2015}, this results 
      in the synthesis between taxonomy and phylogeny.

    For the state information the Global Biotic Interactions database (GloBI) is taken 
      \cite{Poelen2014} (downloaded on 29.01.2018). \\
    This database consists of entries of the form: species A (source) interacts with B (target). \\
    A number of interactions have been identified\footnote{\hyperlink{
      https://github.com/jhpoelen/eol-globi-data/blob/master/eol-globi-lib/src/main/java/org/eol/globi/domain/InteractType.java
      }{https://github.com/jhpoelen/eol-globi-data/.../InteractType.java}; Last checked: 22.03.2018.}
      , including those indicating whether the species source or target has become a parasite or a 
      free-living species from the biological perspective. They are the following:
    \begin{itemize}
      \item free-living source: preysOn, eats, flowersVisitedBy, hasPathogen, pollinatedBy, 
        hasParasite, hostOf
      \item free-living target: preyedUponBy, parasiteOf, visitsFlowersOf, pathogenOf, hasHost
      \item parasite source: parasiteOf, pathogenOf
      \item parasite target: hasParasite, hasPathogen
    \end{itemize}
    Of these interactions, e.g. species A parasitize species B, the state of the species is 
      determined, here is species A parasitic and species B free-living. In the case that a parasite 
      conquers (parasitizes) another parasite yields conflicting states for the second species. 
      This is solved by preferring parasitic. \\
    For each species known identifyers are stored in GloBI. This includes OTT (the taxonomy of OTL). 
      All species that have stored an OTT indentifyer and have a matching interaction are divided into 
      two lists: parasites and free-livings.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Metadata analysis} \label{sec:methods - metadata analysis}
    Based on the input data, generalized linear models are compared with poisson respectively binomial 
      regression according to their residuals. In order to compare models of different complexity, the 
      BIC (Bayesian Information Criterion) values were calculated in addition to the residuals. \\
    There are two different information criteria: AIC (Akaike Information Criteria) and BIC. \\
    The advantage of the BIC is that the penalty is dependent on the sample size and is therefore 
      advantageous for large samples. \\
    For all these calculations, the following R functions were used: \textit{glm()}, \textit{anova()} 
      and \textit{BIC()}.
    % https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html

    For each node, depth, min, max and mean height were noted. The node depth was calculated as the
      distance (number of edges) to the root node. The node height was calculated in three different 
      ways: min, max and mean height refers to the (mininmal, maximal, mean) distance to a leaf node. \\
    The influence in the modeling of these parameters was tested, additively as well as 
      multiplicatively to the models.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Ancestral state reconstruction methods} \label{sec:methods - ancestral state reconstruction}
    The methods used to reconstruct the ancestral states are Fitch parsimony \cite{Fitch1971} and 
      Sankoff parsimony \cite{Sankoff1975}, which are explained in the following subsections.

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Fitch maximum parsimony}
      Based on the work of Cunningham et al. \cite{Cunningham1998} the Fitch method \cite{Fitch1971} 
        was implemented and extended by us for the multifurcated tree case. To understand the 
        differences to the multifurcated case, the algorithm for the binary case is briefly explained 
        and referred to the extension.

      Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (e.g. \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}).

      The algorithm traverses three times through the tree and fills these sets. \\
      In each step, two sets are considered and their intersection is formed. There are two cases:
      \begin{enumerate}
        \item The intersection is not empty and corresponds to the new set.
        \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
      \end{enumerate}
      First traverse from the leaf nodes to the root: each internal node is formed from its child nodes, 
        where the only information lies at the beginning. \\
      Second traverse from the root node to the leafs: each internal node is formed from its parent node 
        and its sibling node. \\
      Last traversion (direction does not matter): build the final state for every node. It is formed 
        from the sets of previous traversals.

      (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
        of internal nodes, so it was just the first traversal.) \\
        \begin{wrapfigure}[20]{r}{0.5\textwidth}
          \centering
          \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
          \caption{Fitch algorithm for binary trees. \\
            The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
            root node) consists of three sets, where the last set is the final one (bold). \\
            From the second internal node (seen from the root node) there are several possibilities to 
            create the second and third set.}
          \label{fig: binary Fitch}
        \end{wrapfigure}
      The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
        more than two children may be present for the first traversal, but the incision or union may 
        also be formed over more than two sets. Also in the second traversing, there may be several 
        sibling nodes. However, there are several possibilities here that were all tested and compared 
        in the simulation. Some of these options are already available in the binary case:
      \begin{itemize}
        \item The parent node has (except for the root node) two state sets, because he came through 
          the up-traversing previously. Are both sets used or only the first traversing?
        \item If there are several siblings, is the cut or union made first of these, or is everything 
          made directly including the parent node.
      \end{itemize}
      The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
        both possibilities of the three sets. \\
      Cunningham uses only the first state set of the parent node \cite{Cunningham1998}. \\
      From these two points four different versions of Fitch were formed:
      \begin{enumerate}
        \item Fitch 1: First state set of parent node; intersection/union of siblings first.
        \item Fitch 2: First state set of parent node; intersection/union of siblings together with parent node.
        \item Fitch 3: Both state sets of parent node; intersection/union of siblings first.
        \item Fitch 4: Both state sets of parent node; intersection/union of siblings together with parent node sets.
      \end{enumerate}

      \begin{figure}[h!]
        \centering
        \includegraphics[trim = 24mm 0mm 28mm 14mm, clip, width=\textwidth]{Figures/simulation_fitch_evaluation.png}
        \caption{Test of Fitch Versions. \todo{beschriften!!}}
        \label{fig:Fitch versions}
      \end{figure}
      These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and
        distributions of 70\% FL to 30\% P and 60\% FL to 40\% P. Figure \ref{fig:Fitch versions} shows 
        this over all unknown node percentage. \\
      % At 90\% unknown nodes and 90\% of multifurcation of the internal nodes, version 1 was 89.26\%, 
      %   version 2 was 89.26\%, version 3 was 89.35\%, and version 4 was 89.31\% correct. Therefore, 
      %   only version 3 was used for all further simulations.
      At 95\% unknown nodes and 95\% of multifurcation of the internal nodes, version 1 was 88.37\%, 
        version 2 was 88.37\%, version 3 was 88.4\%, and version 4 was 88.39\% correct. Therefore, 
        only version 3 was used for all further simulations.

@@

    %---------------------------------------------------------------------------------------------------
    \subsubsection{Sankoff}
      Maximum parsimony algorithm from Sankoff implemented in the R package castor \cite{Louca2017}.
        From this the function \textit{$hsp\_max\_parsimony()$} is used with default settings including
        $transition\_costs="all\_equal"$. \\

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Simulation}\label{sec:methods - simulation}
    The simulation compares these different ancestral state reconstruction algorithms with each 
      other. \\
    First different implementations of the Fitch maximum parsimony are compared and then the best of 
      them is compared with the implementation of the Sankoff algorithm of the \textit{Castor} package 
      \cite{Louca2017}.
